---
title: "ECo 602 - Analysis of Environmental Data"
subtitle: "Final Project"
author: "Liam Waters"
date: "Fall 2020"
output:
  html_document:
    theme: readable
    toc: TRUE
    toc_float: TRUE
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
options(knitr.duplicate.label = TRUE)
```




<!-- The following text won't be displayed in your document.  It tells R how to make nicer looking buttons for your tabbed content. -->

<style type="text/css" rel="stylesheet">

.btn {
    border-width: 0 0 0 0;
    font-weight: normal;
    text-transform: none;
}

.btn-default {
    color: #2ecc71;
    background-color: #ffffff;
    border-color: #ffffff;
}
</style>



# Introduction

This is an overview of most of the R stats features that were covered in ECo 602/634 and how to use them. This page is a work in progress and will likely be updated again in the future to be more comprehensive and perhaps more visually appealing.


# Functions 1: Data Structure Functions {.tabset .tabset-pills}



## `c()`


The function `c()` *combines* or *concatenates* its arguments into a vector (a 1-dimensional data structure consisting of 1 or more elements).

- All of the elements must be of the same *type*.
  - I can't combine character and numeric types in the same call to `c()`

Here's two examples using numeric and character data types:

```{r}
# Create a vector of numbers:
num_vec  = c(1, 4, 8, 9, 13)

# Create a vector of characters:
char_vec = c("a", "fish", "data is cool")
```

I can show the contents of a vector by typing the name of the vector, or using the `print()` function.

```{r}
# Typing the name of the vector into the console prints the contents
num_vec

# The print() function accomplishes the same task:
print(char_vec)
```


## `data.frame()`

A dataframe stores data in a table of variables. The most common way of getting data into a data frame is through the `read.csv()` function, but if you have data that is in another format the `data.frame()` function converts that data into a data frame. The data columns have to have the same number of attributes and the data within each has to be the same type.


To subset a data frame:

- $columnname if the name is known
- The `str()` function can be used to show the names of the columns


```{r data_frame_eg_1}
dat_1 = data.frame(
  letters = c("b", "w", "B"),
  numbers_1 = c(34, 6, 123454)
)

dat_1$letters

str(dat_1)

```


## `matrix()`

Creates a matrix from the given data. `nrow = ` gives the number of rows the data should be divided into. If the data doesn't divide evenly, values are repeated from the start. `byrow = TRUE` can be used to order the data by row instead of columns (default).

```{r matrix_eg_1}
mat_1 = matrix(1:25, nrow = 4)

mat_1
```



## `length()`

Returns the legnth of whatever object was defined. Works on basically all types of values though typically most useful for vectors.

```{r length_eg_1}
num_vec
length(num_vec)

```



## `nrow()`

Returns the number of rows present in an object. 

```{r nrow_eg_1}
nrow(dat_1)

```



## `ncol()`

Returns the number of columns present in an object.

```{r ncol_eg_1}
ncol(dat_1)

```



## `dim()`

Returns the dimensions of an object. The first number is the number of rows and the second number is the number of columns.

```{r dim_eg_1}
dim(mat_1)

```


# Functions 2: Numerical Data Exploration  {.tabset .tabset-pills}

## `summary()`

Summarizes the object and displays the quantiles if applicable.

```{r summary_eg_1}
summary(dat_1)

```


## `mean()`

Returns the mean of a numerical object.

```{r mean_eg_1}
mean(num_vec)
```



## `sd()`
 
Returns the standard deviation of a numerical object. 

```{r sd_eg_1}
sd(num_vec)
```





# Functions 3: Graphical Data Exploration {.tabset .tabset-pills}


## `par()`

`par()` is used to set graphical parameters. For this class `par()` was always used in conjunction with the argument `mfrow = ` and a `c()` vector to display multiple charts on a single image.
        
```{r par_eg_1, fig.asp=1.0}
#creating the data for the top histogram
n = 30
norm_30 = rnorm(n)

#creating the data for the bottom histogram
n = 300
norm_300 = rnorm(n)


par(mfrow = c(2,1)) #the first number is rows and the second number is columns

hist(norm_30, main = "Histogram of random points n = 30")
hist(norm_300, main = "Histogram of random points n = 300")
```

        



## `plot()` 

Used to create scatterplots. There are lots of arguments to customize the output, some of which are shown here.

- `main = ` is the main title. It get bolded and is displayed across the top of the image.
- `sub = ` is the subtitle. It goes on the bottom underneath the x-axis label.
- `xlab = ` and `ylab = ` are the x-axis label and y-axis label respectively.
- `xlim = ` and `ylim = ` take a `c()` with the lower and upper limits for the x-axis and y-axis.
- `cex = ` takes an integer (1 is the default) for the magnification of the plot components.
- `pch = ` sets the shape of the points on the scatterplot. Takes a numeric input.
- `col = ` defines the color of the points, can take a string value as long as it matches one that is in the system. A list of colors can be found at: https://github.com/Liammwaters/ECo_602_2020/blob/d3525acbbf5f4fc3bc82457be59d0539cabe94a0/data/Rcolor.pdf


```{r scatterplot_1, fig.asp=1.0}

#to load a package, use `require()`
require(palmerpenguins)

#formula notation uses the `~`

plot(bill_length_mm ~ body_mass_g, data = penguins,
     main = "penguin bill length as expressed by body mass",
     sub = "this is the subtitle",
     xlab = "body mass in grams",
     ylab = "bill length in mm",
     xlim = c(4200, 6500),
     ylim = c(32, 63),
     cex = 1.5,
     pch = 16,
     col = "green"
)


```



## `hist()`

Used to create histograms. Similar customization arguments as `plot()`. One new one is `breaks = ` which which can be used to set the width of each bar, or column, in the histogram. It can take the form of:
- A single number for the the number of bars (but R treats this as mearly a suggestion and will ignore you if it thinks that will be ugly for some reason)
- A `c()` giving the breakpoints between bars (which R will not ignore) 
       
```{r hist_eg_1, fig.asp=1.0}
#creating the data for the histogram
n = 30
norm_30 = rnorm(n)

#example 1, with breaks = a single numeric suggestion
hist(norm_30, main = "Histogram of random points n = 30", breaks = 17)

#example 2, with breaks = a specific vector
hist(norm_30, main = "Histogram of random points n = 30", breaks = c(-3, -2, -1, 0, 1, 2, 3))
```




## `boxplot()`

Used to create a box-and-whisker plot of the data.
        You must include two examples: a simple boxplot (i.e. one box) and a conditional boxplot (i.e. a single plot that has multiple boxes, conditioned on a categorical variable).

```{r boxplot_eg_1, fig.asp=1.0}

#to load a package, use `require()`
require(palmerpenguins)

#simple boxplot
boxplot(penguins$bill_length_mm, main = "combined penguin bill length", ylab = "bill length in mm")

#conditional boxplot that separates the bill length by the species of penguin
boxplot(
  bill_length_mm ~ species, data = penguins,
  main = "penguin bill length separated by species",
  ylab = "bill length in mm"
)
```




# Functions 4: Distribution Functions {.tabset .tabset-pills}

These are all closely related functions, the first set is for normally distributed data, the second set is for binomially distributed data. Default values: `mean = 0` and `sd = 1`

## `dnorm()`

Gives the value of the probability density function of the normal distribution. Used for continuous data.(the value is taken from the height of the curve at the x-value)

```{r dnorm_eg_1}

#create data values
#let's say that this represents length measurements from fish in a pond
x = seq(0, 20, length.out = 1000)

#use `dnorm()` to show the probability density (y-value on the chart) and set the mean and sd values
y = dnorm(x, mean = 10, sd = 3)

#plot the results
plot(x, y, main = "Normal PDF with mean of 10 and sd = 3", type = "l")
  
```


## `pnorm()`

Gives the value of the cumulative density function of the normal distribution. (the area under the curve to the left of the x-value unless `lower.tail = FALSE` in which case it is the area to the right of x.)

```{r pnorm_eg_1}
#create data
#to find percentage of fish smaller than 12cm in a pond where the mean = 10cm and sd = 3
pnorm(12, 10, 3)
```


## `qnorm()` 

Returns the quantile for a given probability. The inverse of `pnorm()`

```{r qnorm_eg_1}
#create data
#to find the quantile of a fish with the probability of 0.75 of being smaller than 24cm in a pond where the mean = 42cm and sd = 3
qnorm(0.75, 10, 3)
```



## `dbinom()`

Returns the probability of success from a set of trials when each trial has the same probability of success. 

```{r dbinom_eg_1}
# what's the probability of having salamanders under 8 logs in 12 flips? *assume there are never more than one salamander under a log and that they are present under 60% of logs in this area
dbinom(8, 12, 0.60)
```




## `pbinom()`

Returns the probability of a value lower than the target. Again, can be reversed by adding `lower.tail = FALSE` as an argument.

```{r pbinom_eg_1}
# what's the probability that you find salamanders under fewer than 8 logs?
pbinom(8, 12, 0.60)
```


## `qbinom()`

Returns the quantile value for a given probability per trial. 

```{r qbinom_eg_1}

# how many salamanders would you need to find to reach the 90th quantile if you flip 12 logs and theres salamanders under 60% of them?
qbinom(0.90, 12, 0.60)
```



# Functions 5: Other Functions {.tabset .tabset-pills}

These are some of the other useful functions that we learned how to use in class.

## `subset()`

Displays a subset of a an object. Can be used on a vector, matrix, or data frame. Arguments take boolean phrases, but they aren't required. Does need to include the data and a logical expression that indicates either what to keep or what to exclude.

```{r subset_eg_1}
#load data
require(palmerpenguins)

#the penguins data has three species: Adelie, Chinstrap, and Gentoo
summary(penguins$species)

#this subsets the penguin data to include only values associated with the species "Gentoo" and writes it to a new object. The "==" means a logical exact match.
dat_gen = subset(penguins, species == "Gentoo")

#the same command now shows only Gentoo for species
summary(dat_gen$species)
```




## `library()` and `require()`

These have already come up in previous code examples. Both `library()` and `require()` load R packages that have already been installed with `install.packages()` so that they can be accessed for use. `library()` returns an error if the package isn't installed and `require()` returns a warning.

```{r require_eg_1}

require(palmerpenguins)

library(pillar)
```



## `read.csv()`

Used to load data from a .csv file into a data frame.Needs either an exact file path, the relative file path if the .csv is stored in the working directory, or can be used with the `here()` function.


```{r read.csv_eg_1}

#load the "here()" function
require(here)


#this locates the "bird.sta.csv" file that is located in my "data" folder within my R project and assigns it to a new object called "dat_bird"
dat_bird = read.csv(here("data", "bird.sta.csv"))

#head() returns only the first few rows of the object.
head(dat_bird)
```




# Part 2: Data Anaylsis

This section will be a complete data analysis of the delomys data set.

# Data Exploration

## Numerical Exploration
The `shapiro.test()` provides a numeric test of normalicy with the null hypothesis being that the data is normally distributed.
```{r numerical_exploration}

#read in the data
require(here)
dat_del = read.csv(here("data", "delomys.csv"))


#get stats for the body mass column...
summary(dat_del$body_mass)
#...and for the body length
summary(dat_del$body_length)

#test for normalicy in the body mass data
shapiro.test(dat_del$body_mass)
#test for normalicy in the body length data
shapiro.test(dat_del$body_length)


```


## Graphical Exploration {.tabset .tabset-pills}


### scatterplot of body mass and body length


```{r graphical_ex, fig.asp= 0.8}

plot(body_mass ~ body_length, data = dat_del,
     col = "darkslategray4",
     pch = 16)
```


### histogram of body mass

```{r mass_hist_1}
hist(dat_del$body_mass, 
     breaks = 80, # narrows the bars of the histogram so that there is less binning of values
     main = "Histogram of Delomys body mass", 
     xlab = "body mass g")
```



### histogram of body length

```{r length_hist_1}
hist(dat_del$body_length,
     breaks = 80, # narrows the bars of the histogram so that there is less binning of values
     main = "Histogram of Delomys body length", 
     xlab = "body length mm")

```


### boxplot of body mass, conditioned on species

```{r mass_sp_box_1}
boxplot(body_mass ~ binomial, data = dat_del,
        main = "Body mass by species",
        xlab = "species",
        ylab = "body mass g")
```


### boxplot of body mass, conditioned on sex

```{r mass_sex_box_1}
boxplot(body_mass ~ sex, data = dat_del,
        main = "Body mass by sex",
        ylab = "body mass g")
```

### boxplot of body mass, conditioned on both species and sex

```{r mass_sp_sex_box_1}
boxplot(body_mass ~ binomial + sex, data = dat_del,
        main = "Body mass by species and sex",
        xlab = " ",
        ylab = "body mass g",
        names = list("D. dorsalis\n female", 
                     "D. sublineatus\n female", 
                     "D. dorsalis\n male", 
                     "D. sublineatus\n male"))  #the \n inserts a line break so that the name is continued below. In this case it is needed to make the name shorter so that they are all displayed.

```


## Questions {.tabset .tabset-pills}

### 1
Qualitatively describe the relationship between body mass and length.

- Does the relationship seem linear, curved, nonexistent?

*Overall there is a fairly strong linear relationship between body mass and body length. This relationship breaks down at the ends a bit, with individuals that are shorter than ~55mm and longer than ~150mm not folowing the same linear slope. I feel like this isn't uncommon to see with the fringe data in a dataset however.*

  
### 2 
Qualitatively describe the shapes of the histograms.

- Do the data appear normally-distributed? Explain why or why not.

*Both the body mass and body length histograms show data that appears to be mostly normally distributed. The body length data does have some outlying data points that are beyond the expected spread or normal data, but overall  both data sets have a strong peak in the middle and a mostly evenly tapered spread outward from the mean. I'd say that the data can be treated as normal due to the fact that it is just a small fraction of points that don't conform.*

### 3
Using both the histograms and normality tests, do you think the (unconditioned) body masses and body length are normally-distributed?

- Contrast your visual assessment of normality to the results of the numerical normality tests.

*The results from the `shapiro.test()` confirm my visual interpretation of the data being normally distributed. Both the body mass and body length data returned p-values of much less than 0.05.*
    
### 4 
Examine the conditional boxplots. Describe any graphical evidence you see for body mass differences based on species and/or sex.

*The boxplot of body mass by species shows that D. dorsalis is a few grams heavier, on average, than D. sublineatus.The boxplot of body mass by sex shows that males are a few grams heavier, on average, than females. The boxplot of body mass by species and sex shows that the relationship is similar across species and sex.*



## 
